---
title: "Assignment_B1"
author: "Nicholas Mogoi: 30656230"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducible retrieval and procesing multiple biological data formats programmatically using R


This document contains the R code to retrieve data from the NCBI database. The following files are therefore generated:

i. GenBank flatfile (.gb) - with annotated information about the sequence
ii. Sequence (.fasta) - containing the sequence
iii. Metadata (.csv) - summary file, containing the selected descriptive fields
iv. All metadata along with all parsed source qualifiers (.json)

In this assignment, a single protein accession WGL30888.1 is used to retrieve and process the data, in line with demonstrating the ability to access NCBI resources, and generate multiple data formats as above. 


### 1. Setup

```{r}

#loading the required libraries
library(rentrez)
library(xml2)
library(jsonlite)
library(tidyverse)

#defining the accession number
acc <- "WGL30888.1"

```



### 2. Retrieving FASTA sequence
Retrieving the FASTA sequence corresponding to the accession number from the protein database.

```{r}

#fetching FASTA sequence, saving and printing a view
fasta_file <- entrez_fetch(
  db = "protein",
  id = acc,
  rettype = "fasta"
)
write(fasta_file, paste0(acc, ".fasta"))
cat(substr(fasta_file, 1, 200))

```


### 3. Retrieving GenBank flatfile (.gb)

Retrieving the GenBank flat file, saving as .gb, and accessing part of the annotated information.

The GenBank file has annotated information protein domains, details of the organism, and functional annotations.

```{r}

#fetching GenBank file, saving and previewing

gb_datafile <- entrez_fetch(
  db = "protein",
  id = acc,
  rettype = "gb"
)

write(gb_datafile, paste0(acc, ".gb"))
cat(substr(gb_datafile, 1, 700))

```


### 4. Retrieving XML metadata
Retrieving full metadata and preparing it for extraction to CSV, JSON

```{r}
#fetching XML metadata, previewing

xml_data <- entrez_fetch(
  db = "protein",
  id = acc,
  rettype = "xml"
)

write(xml_data, paste0(acc, ".xml"))
cat(substr(xml_data, 1, 600))

```


### 5. Extracting metadata- creating the CSV file
Parsing the XML file to extract key metadata and creating CSV file, for downstream analyses.

```{r}
#extracting metadata summary in CSV format

#parsing XML
xml_parsed <- read_xml(xml_data)

#extracting key metadata fields
accession <- xml_text(xml_find_all(xml_parsed, ".//GBSeq_primary-accession"))
definition <- xml_text(xml_find_all(xml_parsed, ".//GBSeq_definition"))
organism <- xml_text(xml_find_all(xml_parsed, ".//GBSeq_organism"))
length <- xml_text(xml_find_all(xml_parsed, ".//GBSeq_length"))

#creating dataframe
metadata_df <- data.frame(
  Accession = accession,
  Definition = definition,
  Organism = organism,
  Length = length
)

#sneak preview of dataframe
metadata_df

#saving as CSV
write.csv(metadata_df, paste0(acc, ".csv"), row.names = FALSE)

```

### 6. JSON conversion
The complete XML metadata is converted into JSON format. This preserves the full structure of the data, including all nested elements and source qualifiers, making it suitable for computational analysis and data exchange.

```{r}
#converting XML to list
xml_list <- as_list(xml_parsed)

#converting list to JSON
json_data <- toJSON(xml_list, pretty = TRUE, auto_unbox = TRUE)

#saving JSON file
write(json_data, paste0(acc, ".json"))

#previewing JSON file
cat(substr(json_data, 1, 500))

```











