---
title: "multiple_SRA_files_download"
author: "Nick Mogoi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Introduction
This script is designed to retrieve multiple sequencing datasets from the NCBI Sequence Read Archive (SRA) using R. The workflow reads accession IDs from a text file and downloads paired-end FASTQ files using the SRA Toolkit.

# 2. Setup

## 2.1 Set working directory

```{r}

getwd()

#ensure the directory has the .txt accession file
list.files()

```


## 2.2 Load required libraries

```{r}

library(readr)
library(stringr)

```


## 2.3 Read the accessions from the .txt file

```{r}
#read accession IDs and
#view accessions

accessions <- read_lines("sra_accessions.txt")
accessions

#error handling
if (length(accessions) == 0){
  stop("No accessions found")
}

```


## 2.4 Create an output directory

```{r}
#create an output directory for the FASTQ files
dir.create("fastq_files", showWarnings = FALSE)

```


## 2.5 Check fasterq-dump availabiltity

```{r}
system("which fasterq-dump")

#if not available, you've got to install sratools via the terminal first

```


## 2.6 Download SRA data

```{r}

for (acc in accession) {
  message ("Downloadig:", acc)
  
  cmd <- paste(
    "fasterq-dump",
    acc,
    "--split-files",
    "--threads 4",
    "-O fastq_files"
  )
  
  system(cmd)
}
```


## 2.7 Compress FASTQ files

```{r}

system("gzip fastq_files/*.fastq")

```


## 2.8 Verify downloaded files

```{r}

list.files("fastq_files")

```



## Summary

This workflow successfully breaks down how to:

i. read accession IDs from a file
ii. automate SRA downloads in R
iii. generate paired-end FASTQ files
iv. organise outputs into directories

This is a scalable and reproducible retrieval of sequencing data for downstream bioinformatics analysis.


