---
title: "30656230_assignment_B2"
author: "Nicholas_30656230"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

getwd()

```


## Reproducible retrieval of multiple SRA data files programmatically using R

This is an automated batch processing R script essential for initial stages of bioinformatics pipelines.


### Read accession files into R

```{r}

#setup, loading libraries, reading accession file, and viewing accessions

library(tidyverse)
accessions <- read_lines("sra_accessions.txt")
accessions

```


```{r}
#check expected length

length(accessions)

```



### Retrieve SRA data 
This step involves looping through the accessions, calling the SRA toolkit (fasterq-dump) from terminal to R, and downloading the FASTQ files.


```{r}

##Download SRA FASTQ files
#loop through accessions, 
#construct system command,
#and run system command

for (acc in accessions) {
  message("Downloading: ", acc)
  
  cmd <- paste(
    "fasterq-dump",
    acc,
    "--split-files",
    "--threads 4"
  )
  
  system(cmd)
}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
